---
title: "Acquisition and analysis of weather data"
subtitle: "Guatemala City"
author: "Georgios Papadopoulos | "
date: "2026-05-10"
output: github_document
knit: >
  (function(inputFile, encoding) {
    rmarkdown::render(
      inputFile,
      encoding = encoding,
      output_format = "github_document",
      output_file = "README.md"
    )
  })
---

**Reproducible weather data acquisition workflow in R using web scraping, data transformation, and visualization techniques.**

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, 
  message = FALSE,
  warning = FALSE,
  out.width  = "100%",
  fig.align = "center"
  )
```

## Introduction

This project demonstrates a reproducible weather data acquisition workflow in R. Weather data for Guatemala City is scraped from timeanddate.com, cleaned into structured tables, visualized, and interpreted.

The analysis covers short-term forecasts, hourly weather patterns, and annual climate averages.

```{r libraries}
library(tidyverse)
library(rvest)
library(httr)
```

## Methods

The original workflow uses `httr` and `rvest` for web scraping, `tidyverse` for data cleaning and transformation, `stringr` for text extraction, `ggplot2` for visualization and `kableExtra` for formatted tables.

However, the website later introduced restrictions on automated requests which resulted in HTTP 403 errors during rendering. To preserve reproducibility, the source webpages were therefore stored locally as HTML files and parsed with the same scraping workflow.

## 1. 48-hour weather forecast

The weather page is requested programmatically, parsed as HTML, and converted into structured tables. The available tables are then inspected to identify the relevant 48-hour forecast data for further cleaning and analysis.

```{r}
url <- "https://www.timeanddate.com/weather/guatemala/guatemala"
#html <- read_html(GET(url, add_headers(`Accept-Language` = "en")))
html <- read_html("html/forecast_48.html")

tables <- html %>% html_table(fill = TRUE)
cat("The website contains", length(tables), "tables")
```

After inspecting the extracted HTML tables, the third table was identified as the relevant 48-hour forecast table. This table is selected as the source for the following cleaning, restructuring and analysis steps.


### Data cleaning

The extracted table is first restructured to ensure that variables and observations are correctly aligned for subsequent analysis.

```{R}
forecast_48 <- tables[[3]]
```

The final row contains a source footnote rather than forecast data, so it is removed from the table.

```{r}
forecast_48 <- forecast_48[-nrow(forecast_48), ]
```

Weather icons are not captured during scraping, which creates an empty intermediate row in the extracted table.

```{r}
forecast_48[2, -1] <- forecast_48[4, -1]
forecast_48 <- forecast_48[-4, ]
```

The column structure still requires normalization. This is the current state:

```{r, echo=FALSE}
colnames(forecast_48)
```

To create meaningful column names, the **time of day** row is extracted and combined with the corresponding day labels into a single header row.

```{R}
time_row <- as.character(unlist(forecast_48[1, ]))
forecast_48 <- forecast_48[-1, ] 
day_names <- colnames(forecast_48) 

new_names <- paste(day_names, time_row, sep = " ") #combine
new_names[1] <- "Variable"
colnames(forecast_48) <- new_names
```

### Table

```{R}
knitr::kable(
  forecast_48,
  caption = "Forecast for the next 48 hours in Guatemala City"
)
```


### Insights

The 48-hour forecast for Guatemala City shows stable weather conditions with no expected rainfall. Temperatures follow a daily cycle, higher in the afternoon around 26 degrees Celsius and lower at night around 15 degrees Celsius, while humidity increases as temperature decreases. Wind speeds remain light and fairly constant and conditions are mostly cloudy throughout.

## 2. Hourly weather forecast

The same scraping workflow is applied to the hourly forecast page, and the extracted table is stored as `forecast_24` for cleaning, transformation and visualization.

```{r}
url <- "https://www.timeanddate.com/weather/guatemala/guatemala/hourly"
#html <- read_html(GET(url, add_headers("Accept-Language" = "en")))
html <- read_html("html/hourly.html")

tables <- html %>% html_table(fill = TRUE)
forecast_24 <- tables[[1]]
```

### Data cleaning

The first extracted row is reassigned as the column header structure of the table.

```{r}
colnames(forecast_24) <- as.character(unlist(forecast_24[1, ]))
forecast_24 <- forecast_24[-1, ]
```

The footnote contains metadata rather than forecast observations and is therefore removed.

```{r}
forecast_24 <- forecast_24[-nrow(forecast_24), ]
```

Non essential columns are removed to keep only the variables needed.

```{r}
forecast_24 <- forecast_24[, -c(2, 4, 5, 7, 9, 10)]
```

The time variable is standardized to a consistent HH:MM format.

```{r}
forecast_24$Time <- substr(forecast_24$Time, 1, 5)
```

### Table

```{r}
knitr::kable(
  forecast_24,
  caption = "24-hour weather forecast in Guatemala City"
)
```


### Plot

The forecast data is standardized and reshaped into long format to support faceted time series visualization.

```{R}
forecast_long <- forecast_24 %>%
  mutate(
    across(c(Temp, Wind, Humidity), ~ as.numeric(str_extract(.x, "\\d+"))),
    Time = factor(Time, levels = Time)
  ) %>%
  pivot_longer(
    cols = c(Temp, Wind, Humidity),
    names_to = "Variable",
    values_to = "Value"
  ) %>%
  mutate(
    Variable = recode(Variable,
      "Temp" = "Temperature (°C)",
      "Humidity" = "Humidity (%)",
      "Wind" = "Wind (km/h)"
    )
  )

ggplot(forecast_long, aes(x = Time, y = Value, group = 1)) +
  geom_line() +
  facet_wrap(~Variable, scales = "free_y", nrow = 3) +
  labs(
    title = "Weather Forecast for the Next 24 Hours in Guatemala City",
    x = "",
    y = ""
  ) + 
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5, size = 6 ),
        plot.title = element_text(hjust = 0.5))
```

### Insights

Temperature follows a typical diurnal cycle, peaking in the early afternoon at around 27 degrees Celcius and gradually decreasing throughout the evening and night to a minimum of approximately 14 degrees Celcius in the early morning hours. After sunrise, temperatures begin to rise again steadily.

Humidity exhibits an inverse relationship with temperature. During the warmest part of the day, humidity levels are lowest (around 33–38%), while as temperatures decrease in the evening and night, humidity rises significantly, reaching peak levels of over 80% during the early morning hours. This indicates a strong negative correlation between temperature and humidity.

Wind speed remains relatively stable throughout the day, fluctuating within a narrow range of approximately 10 to 14 km/h. Rather than showing a strong upward or downward trend, wind patterns vary in small cycles or phases over the course of the day, with slight increases during the afternoon and late evening.

## 3. Annual weather averages

### Data cleaning

A request is sent to retrieve the climate webpage, which is then parsed into HTML format.

```{R}
url <- "https://www.timeanddate.com/weather/guatemala/guatemala/climate"
#html <- GET(url, add_headers("Accept-Language" = "en")) %>% read_html()
html <- read_html("html/climate.html")
```

Since the climate information is embedded within a specific HTML section, `html_node()` is used to target the climate table directly instead of relying on automatic table extraction.

```{R}
climateTable <- html %>% html_node("#climateTable")
```

The climate section is organized into 13 monthly blocks: one annual summary and 12 individual months. Each block is selected using the `climate-month` class, and `lapply()` is used to extract the month name and weather values. Regular expressions are then applied to isolate numeric values and convert them into a structured data frame.

```{r}
weather_averages <- climateTable %>%
  html_nodes(".climate-month") %>%
  lapply(function(m) {
    
    vals <- m %>% html_nodes("p") %>% html_text(trim = TRUE)
    
    data.frame(
      Month = m %>% 
        html_node("h3") %>% 
        html_text(trim = TRUE) %>% 
        str_extract("^[A-Za-z]+"),
      High  = as.numeric(str_extract(vals[1], "\\d+")),
      Low   = as.numeric(str_extract(vals[2], "\\d+")),
      Mean  = as.numeric(str_extract(vals[3], "\\d+")),
      Precipitation = as.numeric(str_extract(vals[4], "\\d+\\.?\\d*"))
    )
    
  }) %>%
  bind_rows()
```

### Table

```{r}
knitr::kable(
  weather_averages,
  caption = "Annual weather averages in Guatemala City"
)
```

### Plot

```{r}
# data prep
plot_df <- weather_averages[weather_averages$Month != "All", ] 
plot_df$Month <- factor(plot_df$Month,
  levels = c("January","February","March","April",
             "May","June","July","August",
             "September","October","November","December"))
scale_factor <- max(plot_df$Precipitation) / max(plot_df$High)

ggplot(plot_df, aes(x = Month)) +
  # precipitation
  geom_area(aes(y = Precipitation / scale_factor, group = 1),
            fill = "steelblue",
            alpha = 0.3) + 
  # temperature range
  geom_ribbon(aes(ymin = Low, ymax = High, group = 1),
              fill = "tomato",
              alpha = 0.7) +

    geom_line(aes(y = Mean, color = "Mean"), size = 1, group = 1) +

  scale_y_continuous(
    name = "Temperature (°C)",
    limits = c(0, max(plot_df$High) * 1.2),
    sec.axis = sec_axis(~ . * scale_factor, name = "Precipitation (mm)")
  ) +
  scale_color_manual(values = c("Mean" = "firebrick")) +
  scale_x_discrete(labels = month.abb) + 
  labs(title = "Annual Weather Averages in Guatemala City",
       subtitle = "Based on low, mean and high Temperatures collected during 1992–2021",
       caption = "Right side precipitation & Left side temperature",
       x = "") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5),
        plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5),
        plot.caption = element_text(hjust = 0.5),
        axis.title.y = element_text(color = "firebrick"),
        axis.title.y.right = element_text(color = "steelblue"),
        legend.position = "none"
        )
```

### Insights

The annual weather averages for Guatemala City show a relatively stable temperature pattern throughout the year, with an overall mean temperature of 21 degrees Celsius, average highs around 27 degrees Celsius, and lows near 14 degrees Celsius. Temperatures peak slightly during the warmer months of April and May, where mean values reach about 22 degrees Celsius, while cooler conditions occur between November to January with mean temperatures around 19 degrees Celsius. 

In contrast, precipitation varies significantly across the year, meaning that in Guatemala City the weather has a wet and dry season. The driest months are December and January, receiving less than 10 mm of rainfall, whereas the rainy season intensifies from May through October, peaking in September with 186.6 mm of precipitation. Overall, the climate can be characterized as mild in temperature but strongly seasonal in rainfall.